---
title: "Introduction to Working with Dates in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE,message=FALSE)
```



## Introduction

In many data files, the date or time of day will be an important variable. In this introductory tutorial, we will learn some basics on to handles dates.


__A Reminder__: Why do `<date>` objects even matter? Compare the following two plots: one made where the date is in `<chr>` form and the other where date is in its appropriate `<date>` form. 

```{r, appendix = TRUE}
library(tidyverse)
library(lubridate)

animal_crossing <- read_csv("data/animal_crossing_holidays.csv") 

animal_crossing %>%
ggplot(data = ., aes(x = Date1, y = Holiday)) +
  geom_point()

animal_crossing %>% 
  mutate(Date_test_plot = dmy(Date1)) %>%
  ggplot(data = ., aes(x = Date_test_plot, y = Holiday)) +
  geom_point()
```

In which plot does the ordering on the x-axis make more sense?

## Dates with `lubridate`

__Goals__:

* use `lubridate` functions to convert a character variable to a `<date>` variable.
* use `lubridate` functions to extract useful information from a `<date>` variable, including the year, month, day of the week, and day of the year.


### Converting Variables to `<date>` 

The `lubridate` package is built to easily work with `Date` objects and `DateTime` objects. 

To begin, here are a few basic functions `today()`, which  prints today's date, and  `now()` prints today's date and time.


```{r, message = FALSE, appendix = TRUE}
today()
now()
```

There are a number of built-in functions to convert character strings to Dates and Times. 

#### Parsing Dates and Times

  - `ymd()`: Parses dates in the format of "year-month-day" and returns a datetime object.
  - `dmy()`: Parses dates in the format of "day-month-year" and returns a datetime object.
  - `mdy()`: Parses dates in the format of "month-day-year" and returns a datetime object.
  - `hm()`: Parses times in the format of "hour-minute" and returns a time object.
  - `hms()`: Parses times in the format of "hour-minute-second" and returns a datetime object.

Here is a quick example showing what they do.

```{r}
# dates in different formats

d1 <- "2023-04-19" # Also known as 
d2 <- "19-04-2023"
d3 <- "04-19-2023"

ymd(d1)
dmy(d2)
mdy(d3)
```


```{r}
# Parse time in different formats
time_hm <- hm("10:15")
time_hms <- hms("10:15:30")
```

As seen before, these also work on variables within data frames (tibbles).

```{r}
animal_crossing %>%
  mutate(Date1_v2 = dmy(Date1)) %>%
  relocate(Date1_v2)
```


#### `year()`, `month()`, and `mday()` 

The functions `year()`, `month()`, and `mday()` can grab the year, month, and day of the month, respectively, from a `<date>` variable. Like the `forcats` functions, these will almost always be paired with a `mutate()` statement because they will create a new variable.

Notice in the animal crossings data there are a number of variables related to these aspects. Here is how they were created.

```{r}

# starting fresh
animal_crossing2 <- animal_crossing %>% select(Holiday, Date1)

# recreate initial 
animal_crossing2 %>%
  mutate(
    Date = dmy(Date1),
    Month = month(Date),
    Year = year(Date),
    Day = mday(Date),
    Month2 = month(Date, label = TRUE, abbr = FALSE),
    
    # a few extras
    Day_in_year = yday(Date),
    Day_of_week = wday(Date, label = TRUE, abbr = TRUE),
    week_of_year = week(Date)
  )
```


## Using parse_date from the readr package

Another common way to work with dates is to use the `parse_date` (and `parse_date_time` function from `lubridate`). This usually requires us to identify a `format` for the date (or date-time) structure.

For example, if you have a date in the format "2023-04-18 09:19:59", you can use the following format string to parse it using the parse_date() function:

```{r}
date_string <- "2023-04-18 09:19:59"
date <- parse_date(date_string, format = "%Y-%m-%d %H:%M:%S")
date
```


Below is a table of common formats. 


| Format | Description                                       |
|--------|---------------------------------------------------|
| %d     | Day of the month as a number (01-31).     |
| %m     | Month as a number (01-12).                |
| %Y     | Year with century (as a four digit number).            |
| %y     | Year without century (00-99). |
| %H     | Hour (24-hour clock) as a decimal number (00-23). |
| %I     | Hour (12-hour clock) as a decimal number (01-12). |
| %M     | Minute as a decimal number (00-59).               |
| %S     | Second as a decimal number (00-59).               |
| %z     | Time zone offset from UTC (e.g., "-0800").        |
| %Z     | Time zone name.                                   |


## Another Example

We have data on flights originating from New York airports in November 2022. 

```{r}

```



```{r}
url <- "https://raw.githubusercontent.com/iramler/stat234/main/notes/data/ny_airports_nov2022.csv"
ny_airports <- read_csv(url)
```

```{r}
ny_airports %>% mutate(FL_DATE = str_remove(FL_DATE, " 12:00:00 AM")) -> ny_airports
```


First, lets reduce this data into just the four airports in Albany, Buffalo, Rochester, and Syracuse. 

```{r}
airports_to_use = paste( c('Albany', 'Buffalo', 'Rochester', 'Syracuse'), "NY", sep = ", ")

upstate_airports <- 
  ny_airports %>%
    filter(ORIGIN_CITY_NAME %in% airports_to_use)
```

Now convert the FL_DATE variable from a `<chr>` to a `<date>`.

```{r}
upstate_airports <-
  upstate_airports %>%
  mutate(Flight_Date = mdy(FL_DATE))
```

* Calculate the average delay time for each airport.


* Calculate the proportion of flights delayed for each airport.


* Which day of the week has the most flights?


Try out this plot.

```{r}
library(ggTimeSeries)
upstate_airports %>%
  group_by(Flight_Date, ORIGIN_CITY_NAME) %>%
  summarise(avg_daily_delay = mean(DEP_DELAY, na.rm=TRUE)) %>%
  ungroup() %>%
 ggplot_calendar_heatmap("Flight_Date",
                        "avg_daily_delay", 
                        ) + 
  facet_wrap(~ORIGIN_CITY_NAME) +
  theme(legend.position = "top") +
  scale_fill_continuous(low = 'green', high = 'red') +
  labs(y = "Day of Week", y = "Month", fill = "Average Delay (min)") +
  coord_flip()

```


## Another Fun Example

https://trends.google.com/trends/explore?geo=US&hl=en

```{r}
library(gtrendsR)

search_terms <- c("pumpkin spice","cold brew")

mysearch <- gtrends(
  keyword = search_terms, onlyInterest = TRUE,
  #time = "all", # this will let us get the data from 2004 - present 
  time = "2013-04-18 2023-04-18"
)

coffee_df <- mysearch[[1]]

head(coffee_df)
tail(coffee_df)
```


* Make a plot of your Popularity variables through time. 

```{r, echo = FALSE, fig.keep = "none"}
ggplot(data = coffee_df, aes(x = date, 
                                  y = hits,
                                  color = keyword)) +
  geom_line(size = 1.5) +
  scale_color_viridis_d(begin = 0, end = 0.9) +
  theme_bw() +
  labs(x= "Date", y = "Popularity",color = "") +
  theme(legend.position = "bottom")
```



